---
title: "Large datasets"
author: "Anonymous"
date: "2025-03-06"
output: html_document
bibliography: references.bib
nocite: |
  @whitlock2020analysis, @nichols
---

## Initialisation

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
rm(list = ls())
library(Rcompadre)
library(tidyverse)
library(here)
library(phytools)
library(taxize)
library(Rage)
library(popbio)
library(svglite)
library(ggbeeswarm)
source(here("Functions.R"))
```

# introduction

# Compadre

## Acessing Data

```{r}
class <- "Aves"
ClassName <- paste(class, "class")
Comadre <- cdb_fetch("comadre")
ComSub <- subset(Comadre,
                   Class == class & #
                   MatrixTreatment == "Unmanipulated" &
                   MatrixCaptivity == "W")
ComFlag <- cdb_flag(ComSub)
ComSubFlag <- subset(ComFlag,
                      check_NA_A == FALSE &
                      check_ergodic == TRUE)
```

## Data

### Histograms

```{r echo=TRUE, message=FALSE, warning=FALSE}
ComSubFlag$lambda <- unlist(lapply(matA(ComSubFlag), 
                                   popbio::lambda))

ComSubFlag$generation_time <- unlist(lapply(matA(ComSubFlag), 
                                   popbio::generation.time))
GVG <- as_data_frame(ComSubFlag) %>%
  select(lambda, generation_time) %>% 
  mutate(across(everything(), ~ na_if(., Inf))) %>% 
  mutate(across(everything(), ~ na_if(., -Inf))) %>%
  drop_na()

#Examine the distribution of population growth rates:
GrowthHist <- histogram(GVG, GVG$lambda,
                        "Population Growth Rate(位)", 
                        paste("Histogram of Growth Rate Within the ", ClassName)
                        , 0.2, TRUE)

GenHist <- histogram(GVG, GVG$generation_time,
                        "Generation Time (years)", 
                        paste("Histogram of Generation Time Within the ", ClassName)
                        , 1, FALSE)



GrowthHist
GenHist
```

## relationship

```{r echo=TRUE, message=FALSE, warning=FALSE}
GenVGrowth <- scatterplot(GVG, GVG$generation_time, 
                          "Generation Time (years)", 
                          GVG$lambda,
                          "Growth Rate (位)", 
                          paste("Generation time Vs Growth Rate in the ", ClassName)
                          , FALSE)

GenVGrowth
```

### Linear model

normal

```{r}
CGVG <- subset(GVG, generation_time < 50)

GVGMod <- lm(lambda ~ generation_time, GVG)
diagnostic_plots("GVGMod DP.svg", GVGMod)

CGVGMod <- lm(lambda ~ generation_time, CGVG)
diagnostic_plots("CGVGMod DP.svg", CGVGMod)
```

log transformed

```{r}
LGVGMod <- lm(log(lambda) ~ log(generation_time), GVG)
diagnostic_plots("LGVGMod DP.svg", LGVGMod)

LCGVGMod <- lm(log(lambda) ~ log(generation_time), CGVG)
diagnostic_plots("LCGVGMod DP.svg", LCGVGMod)
```

polynomial

```{r}
PGVGMod <- lm(lambda ~ generation_time + I(generation_time^2) + I(generation_time^3), GVG)
diagnostic_plots("PGVGMod DP.svg", PGVGMod)

PCGVGMod <- lm(lambda ~ generation_time + I(generation_time^2) + I(generation_time^3), CGVG)
diagnostic_plots("PCGVGMod DP.svg", PCGVGMod)
```

## Spearman's rank correlation coefficient

-   other things very much not working

-   so trying a non parametric method

-   meets the assumptions

    -   monotonic

    -   outlier doesn't matter as robust

    -   indepedent observations

```{r}
cor.test(GVG$lambda, GVG$generation_time, method = "spearman")
```

-   very significant

-   moderate positive correlation, but doesn't describe the data too well

### Plotting

```{r}
GenVGrowth2 <- scatterplot(GVG, GVG$generation_time, 
                          "Generation Time (years)", 
                          GVG$lambda,
                          "Growth Rate (位)", 
                          paste("Generation time Vs Growth Rate in the", ClassName)
                          , TRUE)
GenVGrowth2
```

## seperating duplicates?

```{r}
ComSingle <- ComSubFlag[which(
  duplicated(ComSubFlag$SpeciesAccepted)==FALSE),]
```

# IUCN RedList

## Acessing data

-   due to issues with data having to load from computer

```{r}
IUCN_data<- read.csv(here("IUCN_comadre_compadre.csv"))

ComIUCN <- ComSubFlag %>%
  left_join(x = ., y = IUCN_data, by = "SpeciesAccepted") %>% 
  mutate(IUCNstatus = case_when(
    IUCNstatus == "EN" ~ "Endangered",
    IUCNstatus == "VU" ~ "Vulnerable",
    IUCNstatus == "NT" ~ "Near Threatened",
    IUCNstatus == "LC" ~ "Least Concern",
    is.na(IUCNstatus) == TRUE ~ "Not Assessed"),
    IUCNstatus = factor(IUCNstatus, 
                        levels = c("Endangered","Vulnerable",
                                   "Near Threatened","Least Concern",
                                   "Not Assessed")))

ComIUCN <- as.data.frame(ComIUCN)

ComIUCN$iucn_colour <- NA
ComIUCN$iucn_colour[which(ComIUCN$IUCNstatus == "Endangered")] <- "Orange" 
ComIUCN$iucn_colour[which(ComIUCN$IUCNstatus == "Vulnerable")] <- "Yellow"
ComIUCN$iucn_colour[which(ComIUCN$IUCNstatus == "Near Threatened")] <- "Dark green"
ComIUCN$iucn_colour[which(ComIUCN$IUCNstatus == "Least Concern")] <- "Green"
ComIUCN$iucn_colour[which(ComIUCN$IUCNstatus == "Not Assessed")] <- "Blue"
```

## analysis

```{r}
IUCNVGVG<- as_data_frame(ComIUCN) %>%
  select(lambda, generation_time, IUCNstatus) %>% 
  mutate(across(c(lambda, generation_time), ~ na_if(., Inf))) %>% 
  mutate(across(c(lambda, generation_time), ~ na_if(., -Inf))) %>%
  drop_na()
```

generation time

```{r}

IUCNVGT <- boxplot(IUCNVGVG, IUCNVGVG$IUCNstatus, "IUCN status", IUCNVGVG$generation_time, "Generation Time (years)", "")
IUCNVGT

ANOVAGT <- aov(generation_time ~ IUCNstatus, data = IUCNVGVG)
summary(ANOVAGT)
```

Growth Rate

```{r}
IUCNVL <- boxplot(IUCNVGVG, IUCNVGVG$IUCNstatus, "IUCN status", IUCNVGVG$lambda, "Growth Rate (位)", "")
IUCNVL

ANOVALambda <- aov(lambda ~ IUCNstatus, data = IUCNVGVG)
summary(ANOVALambda)
```

-   

# open tree of life

```{r echo=TRUE, message=FALSE, warning=FALSE}
ComSingle <- ComSubFlag[which(
  duplicated(ComSubFlag$SpeciesAccepted)==FALSE),]

temp <- as_data_frame(ComSingle)

ResNames <- rotl::tnrs_match_names(names = ComSingle$SpeciesAccepted)

ComSingle$ott_id <- ResNames$ott_id
ComSingle <- ComSingle[-which(is.na(ComSingle$ott_id)),]

temp2 <- as_data_frame(ComSingle)

ResNames <- rotl::tnrs_match_names(names = ComSingle$SpeciesAccepted)

ComSingle$OTL_unique_name <- ResNames$unique_name
temp3 <- as_data_frame(ComSingle)
ComSingle <- as_data_frame(ComSingle)
row.names(ComSingle) <- ResNames$unique_name

temp4 <- as_data_frame(ComSingle)

```

```{r}

#This tree contains information for most of the species in COMADRE and COMPADRE, and can also be found here https://github.com/jonesor/compadreDB/tree/master/ under the subfolder "Phylogeny" 

tree <- read.tree("COMPADRE-COMADRE_Phylo_June_16_2019.tre")

#Let's clean up the names in the tree, so we can compare them with the names in our object
tree$tip.label <- gsub("_", " ", tree$tip.label)

ComSingle <- ComSingle  %>% 
  mutate(across(c(lambda, generation_time), ~ na_if(., Inf))) %>% 
  mutate(across(c(lambda, generation_time), ~ na_if(., -Inf))) %>%
  drop_na(lambda, generation_time)

#Now we need to "prune" the tree, which means to cut out the branches that are not present in our data:
PrunedTree <- drop.tip(tree, setdiff(tree$tip.label, ComSingle$OTL_unique_name))

#The resulting tree has 34 tips, meaning 34 speci

#So naturally it seems like we are missing some from our object, which are not present in the tree.

ComSingle <- ComSingle[match(PrunedTree$tip.label,ComSingle$OTL_unique_name),]
row.names(ComSingle) <- ComSingle$OTL_unique_name
#Checking that the order is the same between the object and the tree

# Let us now plot the values of population growth rate at the tips of the phylogeny, and reconstruct the ancestral state of the originating lineange
log_lambdas <- log(setNames(ComSingle$lambda, row.names(ComSingle)))
# create "contMap" object, which allows for the reconstruction of likely ancestral traits



Aves_contMap_lambda <- contMap(PrunedTree, log_lambdas,plot=FALSE,res=1000, method="anc.ML")
Aves_contMap_lambda <- setMap(Aves_contMap_lambda, c("white","#FFFFB2","#FECC5C","#FD8D3C","#E31A1C"))
plot(Aves_contMap_lambda, fsize=c(0.7,0.8), leg.txt="log(population growth rate)")
par(mar=c(5.1,4.1,4.1,2.1)) ## reset margins to default

#Let's do the same thing for generation time
log_gen_T <- log(setNames(ComSingle$generation_time, row.names(ComSingle)))

Aves_contMap_gen_T <- contMap(PrunedTree, log_gen_T,plot=FALSE,res=200, method="anc.ML")
Aves_contMap_gent_T <- setMap(Aves_contMap_gen_T, c("white","light blue","blue","violet","purple"))
plot(Aves_contMap_gent_T, fsize=c(0.7,0.8), leg.txt="log(generation time) (years)")
```

## population performance

using code from [@exercise]

```{r}
## creating new column
PopPerf <- 
  as_data_frame(ComSingle$lambda) %>% 
  mutate(category = factor(
    case_when(value < 1  ~ "Decrease",
              value > 1  ~ "Increase")
  ))

row.names(PopPerf) <- row.names(ComSingle)
PopPerf <- as.factor(setNames(PopPerf$category,rownames(PopPerf)))

PerfTree <- dotTree(PrunedTree, PopPerf, colors=setNames(c("#EA0000", "#1E88E5"), c("Decrease","Increase")), ftype="i", fsize=0.3)


```
